Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-05 formally.tex
tweaked colors
tweaks

The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-expressions} and \textbf{H-types} (Sec. \ref{sec:holes}), i.e. expressions and types with \emph{holes}. H-types classify H-expressions according to Hazelnut's \textbf{bidirectionally typed static semantics}.
\item \textbf{Z-expressions} and \textbf{Z-types} (Sec. \ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-expressions and H-types, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-expression (resp. Z-type) corresponds to an H-expression (resp. H-type) by \emph{cursor erasure}.
\item \textbf{Actions} (Sec. \ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectionally typed action semantics}. The action semantics enjoys a rich metatheory. Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
\end{itemize}

@@

\subsection{Holes}\label{sec:holes}

@@

Rule (\ref{rule:syn-num}) states that number literals synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.

The rules given so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.

@@

A non-empty hole contains an H-expression that is ``under construction'', as described in Sec. \ref{sec:example}. According to Rule (\ref{rule:syn-hole}), the  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type.

@@

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
\label{fig:zexp-syntax}
\end{figure}

>>>>

The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are types and expressions with {holes}. H-types classify H-expressions according to Hazelnut's \textbf{bidirectional static semantics}.
\item \textbf{Z-expressions} and \textbf{Z-types} (Sec. \ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-expressions and H-types, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-expression (resp. Z-type) corresponds to an H-expression (resp. H-type) by \emph{cursor erasure}.
\item \textbf{Actions} (Sec. \ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectional action semantics}. The action semantics enjoys a rich metatheory. Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
\end{itemize}

@@

\subsection{H-types and H-expressions}\label{sec:holes}

@@

Rule (\ref{rule:syn-num}) states that numbers synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.

The rules described so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.

@@

A non-empty hole contains an H-expression that is ``under construction'', as described in Sec. \ref{sec:example}. According to Rule (\ref{rule:syn-hole}), this inner  expression must synthesize some type, but, like the empty hole,  non-empty holes synthesize the type hole.

@@

\end{subequations}
\subsection{Z-expressions and Z-types}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
\label{fig:zexp-syntax}
\end{figure}
